Tail associations and their impact on extinction risk

### 2019-11-09 MainText.Rmd
fixed typos

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across all or most habitat patches. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods (these studies used Pearson correlations).
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, including how such 
associations may be changing. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be and appropriate and important quantity to study 
for many applications. For instance, in cases for which $X_i$ has a linear 
or approximately linear influence on some aspect of the ecosystem
at location $i$, Pearson correlations of the ecological impacts at locations 
$i$ and $j$, which may be of fundamental interest, will relate to 
Pearson correlations between $X_i$ and $X_j$. But in many cases the
ecological impact of $X_i$ and $X_j$ will be a nonlinear or unknown monotonic functions of 
the values of $X_i$ and $X_j$. So studying rank-based correlations
may be more informative, ecologically as well as providing the mathematical advantages
listed above.

@@

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
approximation of the long-term stochastic growth rate.
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered, 
if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using the well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems worth explicitly exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates
and relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

>>>>

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the opposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across all or most habitat patches. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods (these studies used Pearson correlations).
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, including how such 
associations may be changing. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be an appropriate and important quantity to study 
for many applications. For instance, in cases for which $X_i$ has a linear 
or approximately linear influence on some aspect of the ecosystem
at location $i$, Pearson correlations of the ecological impacts at locations 
$i$ and $j$, which may be of fundamental interest, will relate to 
Pearson correlations between $X_i$ and $X_j$. But in many cases, the
ecological impact of $X_i$ and $X_j$ will be a nonlinear or unknown monotonic function of 
the values of $X_i$ and $X_j$. So studying rank-based correlations
may be more informative, ecologically as well as providing the mathematical advantages
listed above.

@@

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
approximation of the long-term stochastic growth rate.
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered, 
if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using the well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extinction risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems worth explicitly exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates
and relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.
